---
title: "Reproducible Research Assignement 2"
author: "TK"
date: "9 août 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(rmarkdown)
library(knitr)
Sys.setlocale("LC_TIME", "English")
setwd("E:/Coursera/Reproducible Research/Week 4/Course Project 2")
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

Storm Data [47Mb]
There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation
National Climatic Data Center Storm Events FAQ
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## 1.Download the file and put the file in the data folder 
```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl,destfile="./data/StormData.csv.bz2")
```
## 2.Read the file

```{r}

Stormdata<-read.csv(bzfile("StormData.csv.bz2"), header=TRUE, stringsAsFactors=FALSE) 

str(Stormdata)

dim(Stormdata)

names(Stormdata)

```

## 3. Selecting Variables

 The dataset has 37 variables. For this analysis we need seven of these variables. These are:
 EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.

 We selecting this variables.

```{r}
variables<-c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")
data<-Stormdata[variables]

dim(data)

names(data)

```

## 4. Selecting events that cause the most fatalities and injuries.
```{r}
fatal <- aggregate(FATALITIES ~ EVTYPE, data = data, sum)
fatal <- fatal [order(fatal$FATALITIES, decreasing=TRUE),]
head(fatal)
```

## 5. Plot Fatalities
```{r}
par(mar=c(12, 6, 1, 1))
barplot (height = fatal$FATALITIES[1:10], names.arg = fatal$EVTYPE[1:10], las = 2, cex.names= 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Fatalities: Top 10 Events", line=-5)
title (ylab = "Total number of Fatalities", line=4)
```

## 6. Selecting events that cause the most injuries.
```{r}
injury <- aggregate(INJURIES ~ EVTYPE, data = data, sum)
injury <- injury[order(injury$INJURIES, decreasing=TRUE),]
head(injury)
```

## 7. Plot Injuries
```{r}
par(mar=c(12, 6, 1, 1))
barplot (height = injury$INJURIES[1:10], names.arg = injury$EVTYPE[1:10], las = 2, cex.names = 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Injuries: Top 10 Events", line=-5)
title (ylab = "Total number of Injuries", line=4)
```

## 8. Total economic damage produced by each type of event.

 Property and crop damages are summed-up over the years from 1950. 

 Finding property damage
 Property damage exponents for each level was listed out and assigned those values for the property exponent   data. 
 Invalid data was excluded by assigning the value as '0'. 
 Then property damage value was calculated by multiplying the property damage and property exponent value.
```{r}
unique(data$PROPDMGEXP)


# Assigning values for the property exponent data 
data$PROPEXP[data$PROPDMGEXP == "K"] <- 1000
data$PROPEXP[data$PROPDMGEXP == "M"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == ""] <- 1
data$PROPEXP[data$PROPDMGEXP == "B"] <- 1e+09
data$PROPEXP[data$PROPDMGEXP == "m"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == "0"] <- 1
data$PROPEXP[data$PROPDMGEXP == "5"] <- 1e+05
data$PROPEXP[data$PROPDMGEXP == "6"] <- 1e+06
data$PROPEXP[data$PROPDMGEXP == "4"] <- 10000
data$PROPEXP[data$PROPDMGEXP == "2"] <- 100
data$PROPEXP[data$PROPDMGEXP == "3"] <- 1000
data$PROPEXP[data$PROPDMGEXP == "h"] <- 100
data$PROPEXP[data$PROPDMGEXP == "7"] <- 1e+07
data$PROPEXP[data$PROPDMGEXP == "H"] <- 100
data$PROPEXP[data$PROPDMGEXP == "1"] <- 10
data$PROPEXP[data$PROPDMGEXP == "8"] <- 1e+08

# Assigning '0' to invalid exponent data
data$PROPEXP[data$PROPDMGEXP == "+"] <- 0
data$PROPEXP[data$PROPDMGEXP == "-"] <- 0
data$PROPEXP[data$PROPDMGEXP == "?"] <- 0

# Calculating the property damage value
data$PROPDMGVAL <- data$PROPDMG * data$PROPEXP
data$PROPDMGVAL <-as.numeric(data$PROPDMGVAL)
```

## Finding crop damage

 Crop damage exponents for each level was listed out and assigned those values for the crop exponent data. 
 Invalid data was excluded by assigning the value as '0'. 
 Then crop damage value was calculated by multiplying the crop damage and crop exponent value.
```{r}
unique(data$CROPDMGEXP)


# Assigning values for the crop exponent data 
data$CROPEXP[data$CROPDMGEXP == "M"] <- 1e+06
data$CROPEXP[data$CROPDMGEXP == "K"] <- 1000
data$CROPEXP[data$CROPDMGEXP == "m"] <- 1e+06
data$CROPEXP[data$CROPDMGEXP == "B"] <- 1e+09
data$CROPEXP[data$CROPDMGEXP == "0"] <- 1
data$CROPEXP[data$CROPDMGEXP == "k"] <- 1000
data$CROPEXP[data$CROPDMGEXP == "2"] <- 100
data$CROPEXP[data$CROPDMGEXP == ""] <- 1

# Assigning '0' to invalid exponent data
data$CROPEXP[data$CROPDMGEXP == "?"] <- 0


# calculating the crop damage 
data$CROPDMGVAL <- data$CROPDMG * data$CROPEXP
data$CROPDMGVAL <-as.numeric(data$CROPDMGVAL)
```

## Total damages was estimated:
```{r}
data$dmgtot <- data$CROPDMGVAL+data$PROPDMGVAL
dmgtot <- aggregate(dmgtot ~ EVTYPE, data, FUN = sum)
dmgtot <- dmgtot[order(dmgtot$dmgtot, decreasing=TRUE),]
dmgtot
barplot (height = dmgtot$dmgtot[1:10]/(10^9), names.arg = dmgtot$EVTYPE[1:10], las = 2, cex.names= 0.8,
         col = rainbow (10, start=0, end=0.5))
title (main = "Total Damages", line=-5)
title (ylab = "Damage Cost ($ billions)", line=4)
```

## Results and Conclusions

 Tornados caused the maximum number of fatalities and injuries. 
 It was followed by Excessive Heat for fatalities and Thunderstorm wind for injuries.
 Floods caused the maximum property damage where as Drought caused the maximum crop damage. 
 Second major events that caused the maximum damage was Hurricanes/Typhoos for property damage and Floods for  crop damage.



